### 2017-08-11 whitepaper.tex
Initial commit, stub out Bioinformatics template

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Phylogenetics}

\title[Nextstrain]{Nextstrain: real-time tracking of virus evolution}
\author[Hadfield \textit{et~al}.]{
James Hadfield\,$^{\text{\sfb 1,}\ast}$,
Colin Megill\,$^{\text{\sfb 1}}$,
Sidney Bell\,$^{\text{\sfb 1,2}}$,
John Huddleston\,$^{\text{\sfb 1,2}}$,
Barney Potter\,$^{\text{\sfb 1}}$,
Charlton Callender\,$^{\text{\sfb 1}}$,
Pavel Sagulenko\,$^{\text{\sfb 3}}$,
Trevor Bedford\,$^{\text{\sfb 1,}\S}$ and
Richard Neher\,$^{\text{\sfb 3,}\S}$}
\address{
$^{\text{\sf 1}}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA,
$^{\text{\sf 2}}$Molecular and Cellular Biology Program, University of Washington, Seattle, WA, USA and
$^{\text{\sf 3}}$Biozentrum, University of Basel, Basel, Switzerland.
}

\corresp{$^\ast$To whom correspondence should be addressed. $^\S$These authors contributed equally to this work.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Summary:} Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text
Text Text Text Text Text.\\
\textbf{Availability and implementation:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
}

\maketitle

We'll definitely cite Nextflu \citep{neher2015nextflu}.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text. Figure~\ref{nextstrain}
shows that the above method  Text Text Text Text  Text Text Text
Text Text Text  Text Text.  \citep{neher2015nextflu} wants to know about
{\ldots}{\ldots} text follows.

\begin{equation}
\sum \text{\it x}+ \text{\it y} =\text{\it Z}
\label{eq:01}
\end{equation}

Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text
text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method Text
Text Text Text Text Text Text Text Text Text Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method Text
Text Text Text Text Text Text Text Text Text Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\subsection*{This is a subheading. These are allowed in an Application Note.}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text\break
text text

\begin{table}[!t]
\processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule head1 &
head2 & head3 & head4\\\midrule
row1 & row1 & row1 & row1\\
row2 & row2 & row2 & row2\\
row3 & row3 & row3 & row3\\
row4 & row4 & row4 & row4\\\botrule
\end{tabular}}{This is a footnote}
\end{table}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.45\textwidth]{figures/nextstrain}}
\caption{Caption, caption.}\label{nextstrain}
\end{figure}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text

Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text
Text Text Text Text. Figure~2\vphantom{\ref{nextstrain}} shows that
the above method Text Text. Text Text Text  Text Text Text Text
Text Text. Figure~2\vphantom{\ref{nextstrain}} shows that the above
method Text Text. Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method Text
Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about  text text text text

Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the
above method  Text Text Text Text Text Text Text Text Text Text
Text Text.  \citealp{neher2015nextflu} might want to know about text
text text text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{neher2015nextflu} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text\break Text.

\vspace*{-10pt}

\section*{Acknowledgements}

Text Text Text Text Text Text  Text Text.  \citealp{neher2015nextflu} might want to know about  text
text text text

\vspace*{-12pt}

\section*{Funding}

This work has been supported by the... Text Text  Text Text.

\vspace*{-12pt}

\bibliographystyle{bioinfo}
\bibliography{whitepaper}

\end{document}

### 2017-08-29 whitepaper.tex
v1

@@
 
-\abstract{\textbf{Summary:} Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text
-Text Text Text Text Text.\\
-\textbf{Availability and implementation:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
-\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
+\abstract{
+\textbf{Summary:} 
+Understanding the spread and evolution of viral pathogens is important for effective public health measures and surveillance. 
+Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases. 
+Nextstrain is a bioinformatics and visualisation platform which presents a real-time view into the evolution and spread of a range of viral pathogens of high human importance. 
+Nextstrain makes these results, representing our current understanding, publicly available for use by health professionals, epidemiologists, virologists and the public alike. 
+\\
+\textbf{Availability and implementation:}
+All code (predominantly Javascript and Python) is freely available from https://github.com/blab/nextstrain and the web-application is available at www.nextstrain.org.
+\\
+\textbf{Contact:} 
+\href{jhadfiel@fredhutch.org}{jhadfiel@fredhutch.org}, \href{tbedford@fredhutch.org}{tbedford@fredhutch.org}, \href{richard.neher@tuebingen.mpg.de}{richard.neher@tuebingen.mpg.de}
+\\
 }
@@
 
-We'll definitely cite Nextflu \citep{neher2015nextflu}.
+Viral pathogens pose an ongoing and ever present danger to human health on a worldwide scale, including epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas, as well as ongoing disease spread such as seasonal influenza.
+The rapid evolution of these viruses allows inference of the evolution and epidemiology of the disease from genomic data. 
+Such analysis are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{Pybus:2013bj}.
+Furthermore, the results of analysis are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding. 
+We have developed nextstrain to visualise viral outbreaks in as close to real time as possible.
+This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text. Figure~\ref{nextstrain}
-shows that the above method  Text Text Text Text  Text Text Text
-Text Text Text  Text Text.  \citep{neher2015nextflu} wants to know about
-{\ldots}{\ldots} text follows.
+The regularly updated nature and rapidity of these analysis is crucial to the monitoring and understanding of pathogen epidemiology and evolution.  
+Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{Quick:2016kf,Faria105171}.
+This rapidity of sequencing must be complemented with rapid methods by which to analyse and interpret results.
 
-\begin{equation}
-\sum \text{\it x}+ \text{\it y} =\text{\it Z}
-\label{eq:01}
-\end{equation}
-
-Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text
-text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method Text
-Text Text Text Text Text Text Text Text Text Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method Text
-Text Text Text Text Text Text Text Text Text Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\subsection*{This is a subheading. These are allowed in an Application Note.}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text\break
-text text
-
-\begin{table}[!t]
-\processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule head1 &
-head2 & head3 & head4\\\midrule
-row1 & row1 & row1 & row1\\
-row2 & row2 & row2 & row2\\
-row3 & row3 & row3 & row3\\
-row4 & row4 & row4 & row4\\\botrule
-\end{tabular}}{This is a footnote}
-\end{table}
+Nextstrain consists of three components:
+``Fauna'' is a collection of python scripts to maintain a database representative of all publicly available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data.
+``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
+This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{Sagulenko:2017vm}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours. 
+These scripts are designed in a manner to be readily co-opted to different pathogens.
+Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
+This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
+There is a growing need for surveillance of non-influenza viruses \citep{FRCPath:2017fg}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.
 
@@
 \centerline{\includegraphics[width=0.45\textwidth]{figures/nextstrain}}
-\caption{Caption, caption.}\label{nextstrain}
+\caption{Currnetly known genomic epidemiology of Zika virus (\href{nextstrain.org/zika}{nextstrain.org/zika}).
+The main interface consists of three linked panels - a phylogenetic tree, the inferred transmissions projected onto a map, and the genetic diversity across the genome.}
+\label{nextstrain}
 \end{figure}
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
+\subsection*{Joint Temporal \& Spatial Visualisation}
+Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
+Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
+These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional metadadata such as genotype or serotype expressed through colourings (figure \ref{nextstrain}).
+This is coupled with a user-interactable time slider to see how the pathogen has evolved and spread over the course of the epidemic.
+By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
+This approach both conveys the geographical spread of disease and the chaotic nature of epidemics.
 
-Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text
-Text Text Text Text. Figure~2\vphantom{\ref{nextstrain}} shows that
-the above method Text Text. Text Text Text  Text Text Text Text
-Text Text. Figure~2\vphantom{\ref{nextstrain}} shows that the above
-method Text Text. Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method Text
-Text.
+\subsection*{Transmission Chain Analysis}
+Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probably transmission events given the sampled data, together with a marginal likelihood for each event.
+Inferred transmissions are displayed as colours on the tree, with opacity used to convey confidence, and each transmission is also drawn on the map. 
+More?
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about  text text text text
+\subsection*{Monitoring of viral evolution \& adaptation}
+Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar-chart of entropy at each position in across the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
+As with all views into data, selecting a position in the genome reveals the selected segregating variant in the phylogeny and the map. 
+This allows interrogation of genetic change which may be adaptive or underlying a change in disease dynamics.
 
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the
-above method  Text Text Text Text Text Text Text Text Text Text
-Text Text.  \citealp{neher2015nextflu} might want to know about text
-text text text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{nextstrain}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{neher2015nextflu} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text\break Text.
+For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
+China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
+The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{doi:10.1056/NEJMoa1304617}.
+\textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{deVries:2017dj}.
+Nextstrain allows monitoring of such mutations, such that interventions and control measures may be implemented where appropriate.
+
+\subsection*{A Model for Public Sharing of Data}
+Nextstrain presents a single overview of both ongoing viral disease (seasonal \& avian influenza, zika) as well as retrospective analysis (dengue, ebola), all based upon the same underlying bioinformatics architecture.
+As such, it is well positioned to respond to both the next outbreak and continued surveillance of current ones. 
+Analysis of such outbreaks relies on public sharing of data, and nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
+Scientists are justifiably skeptical of losing control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
+We do, however provide the ability to download phylogenetic trees, metadata and screenshots, and private metadata can be added by users by simply dragging a CSV file onto the browser. 
+We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, and raises the possibility of further collaboration between scientists.
+Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{Croucher:2015go}.
 
@@
 \section*{Acknowledgements}
-
-Text Text Text Text Text Text  Text Text.  \citealp{neher2015nextflu} might want to know about  text
-text text text
+We would like to thank...
 
@@
 
-This work has been supported by the... Text Text  Text Text.
+This work has been supported by the...
 

### 2017-09-06 whitepaper.tex
Fix references

@@
 The rapid evolution of these viruses allows inference of the evolution and epidemiology of the disease from genomic data.
-Such analysis are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{Pybus:2013bj}.
+Such analysis are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
 Furthermore, the results of analysis are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding.
@@
 The regularly updated nature and rapidity of these analysis is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
-Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{Quick:2016kf,Faria105171}.
+Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
 This rapidity of sequencing must be complemented with rapid methods by which to analyse and interpret results.
@@
 ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
-This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{Sagulenko:2017vm}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours. 
+This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
 These scripts are designed in a manner to be readily co-opted to different pathogens.
@@
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
-There is a growing need for surveillance of non-influenza viruses \citep{FRCPath:2017fg}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.
+There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.
 
@@
 China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
-The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{doi:10.1056/NEJMoa1304617}.
-\textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{deVries:2017dj}.
+The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
+\textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
 Nextstrain allows monitoring of such mutations, such that interventions and control measures may be implemented where appropriate.
@@
 We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, and raises the possibility of further collaboration between scientists.
-Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{Croucher:2015go}.
+Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{croucher2015application}.
 

### 2017-09-07 whitepaper.tex
small edits

@@
 Trevor Bedford\,$^{\text{\sfb 1,}\S}$ and
-Richard Neher\,$^{\text{\sfb 3,}\S}$}
+Richard A.~Neher\,$^{\text{\sfb 3,4,5}\S}$
+}
 \address{
@@
 $^{\text{\sf 2}}$Molecular and Cellular Biology Program, University of Washington, Seattle, WA, USA and
-$^{\text{\sf 3}}$Biozentrum, University of Basel, Basel, Switzerland.
+$^{\text{\sf 3}}$Max Planck Institute for Developmental Biology, T\"ubingen, Germany.
+$^{\text{\sf 4}}$Biozentrum, University of Basel, Basel, Switzerland.
+$^{\text{\sf 5}}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland.
 }
@@
 \textbf{Availability and implementation:}
-All code (predominantly Javascript and Python) is freely available from https://github.com/blab/nextstrain and the web-application is available at www.nextstrain.org.
+All code (predominantly Javascript and Python) is freely available from https://github.com/nextstrain and the web-application is available at www.nextstrain.org.
 \\
 \textbf{Contact:}
-\href{jhadfiel@fredhutch.org}{jhadfiel@fredhutch.org}, \href{tbedford@fredhutch.org}{tbedford@fredhutch.org}, \href{richard.neher@tuebingen.mpg.de}{richard.neher@tuebingen.mpg.de}
+\href{jhadfiel@fredhutch.org}{jhadfiel@fredhutch.org}, \href{tbedford@fredhutch.org}{tbedford@fredhutch.org}, \href{richard.neher@unibas.ch}{richard.neher@unibas.ch}
 \\
@@
 We have developed nextstrain to visualise viral outbreaks in as close to real time as possible.
-This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
+%This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
 
@@
 Nextstrain consists of three components:
-``Fauna'' is a collection of python scripts to maintain a database representative of all publicly available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data.
+``Fauna'' is a collection of python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data.
 ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
 This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
-These scripts are designed in a manner to be readily co-opted to different pathogens.
+These scripts separate generic core functionality from a light virus specific layer such that they are easily co-opted to different pathogens.
 Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
@@
 \centerline{\includegraphics[width=0.45\textwidth]{figures/nextstrain}}
-\caption{Currnetly known genomic epidemiology of Zika virus (\href{nextstrain.org/zika}{nextstrain.org/zika}).
+\caption{Currently known genomic epidemiology of Zika virus (\href{nextstrain.org/zika}{nextstrain.org/zika}).
 The main interface consists of three linked panels - a phylogenetic tree, the inferred transmissions projected onto a map, and the genetic diversity across the genome.}
@@
 This work has been supported by the...
+OpenSciencePrize to TB and RAN, the ERC through StG-260686 to RAN.
 

### 2017-10-09 whitepaper.tex
minor improvements

@@
 Nextstrain consists of three components:
-``Fauna'' is a collection of python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data.
+``Fauna'' is a collection of python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, GISAID and ViPR, as well as github repositories and other sources of genomic data.
 ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
@@
 By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
-This approach both conveys the geographical spread of disease and the chaotic nature of epidemics.
+This approach both communicates the geographical spread of disease and the chaotic nature of epidemics.
 
@@
 Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probably transmission events given the sampled data, together with a marginal likelihood for each event.
-Inferred transmissions are displayed as colours on the tree, with opacity used to convey confidence, and each transmission is also drawn on the map.
-More?
+Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
+It is important to make known the confidence of any such reconstruction, and we do this in two ways: by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
 
@@
 
-\section*{Acknowledgements}
-We would like to thank...
-
-\vspace*{-12pt}
+% please add acknowledgment if applicable.
+%\section*{Acknowledgements}
+%We would like to thank...
+%
+%\vspace*{-12pt}
 
@@
 
-This work has been supported by the...
-OpenSciencePrize to TB and RAN, the ERC through StG-260686 to RAN.
+This work has been supported by the OpenSciencePrize to TB and RAN and the ERC through StG-260686 to RAN.
 

### 2017-10-09 whitepaper.tex
Edit SMB author + funding info

@@
 Colin Megill\,$^{\text{\sfb 1}}$,
-Sidney Bell\,$^{\text{\sfb 1,2}}$,
+Sidney M. Bell\,$^{\text{\sfb 1,2}}$,
 John Huddleston\,$^{\text{\sfb 1,2}}$,
@@
 
-This work has been supported by the OpenSciencePrize to TB and RAN and the ERC through StG-260686 to RAN.
+This work has been supported by the OpenSciencePrize to TB and RAN, the ERC through StG-260686 to RAN, and the NSF through DGE-1256082 to SMB.
 

### 2017-10-09 whitepaper.tex
Reclassify dengue (it is an ongoing/endemic pathogen)

@@
 \subsection*{A Model for Public Sharing of Data}
-Nextstrain presents a single overview of both ongoing viral disease (seasonal \& avian influenza, zika) as well as retrospective analysis (dengue, ebola), all based upon the same underlying bioinformatics architecture.
+Nextstrain presents a single overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, zika, ebola), all based upon the same underlying bioinformatics architecture.
 As such, it is well positioned to respond to both the next outbreak and continued surveillance of current ones.

### 2017-10-09 whitepaper.tex
Feel free to take 'em or leave 'em: minor changes to word choice  / a couple pesky typos.

@@
 The rapid evolution of these viruses allows inference of the evolution and epidemiology of the disease from genomic data.
-Such analysis are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
-Furthermore, the results of analysis are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding.
+Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
+Furthermore, the results of analyses are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding.
 We have developed nextstrain to visualise viral outbreaks in as close to real time as possible.
-%This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
+% This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
 
-The regularly updated nature and rapidity of these analysis is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
+The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
 Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
@@
 ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
-This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
+This leverages the maximum likelihood phylodynamic analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
 These scripts separate generic core functionality from a light virus specific layer such that they are easily co-opted to different pathogens.
@@
 By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
-This approach both communicates the geographical spread of disease and the chaotic nature of epidemics.
+This approach both communicates the geographical spread of disease and the stochastic nature of epidemics.
 
 \subsection*{Transmission Chain Analysis}
-Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probably transmission events given the sampled data, together with a marginal likelihood for each event.
+Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
 Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
@@
 Analysis of such outbreaks relies on public sharing of data, and nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
-Scientists are justifiably skeptical of losing control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
+Scientists are justifiably hesitant to share control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
 We do, however provide the ability to download phylogenetic trees, metadata and screenshots, and private metadata can be added by users by simply dragging a CSV file onto the browser.
-We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, and raises the possibility of further collaboration between scientists.
+We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.
 Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{croucher2015application}.

### 2017-10-10 whitepaper.tex
tweaks + change in abstract

@@
 Understanding the spread and evolution of viral pathogens is important for effective public health measures and surveillance.
-Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
-Nextstrain is a bioinformatics and visualisation platform which presents a real-time view into the evolution and spread of a range of viral pathogens of high human importance.
+%Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
+Nextstrain consists of a bioinformatics pipeline for phylodynamics analysis of viral genomes and a to an interactive visualisation platform which presents a real-time view into the evolution and spread of a range of viral pathogens of high human importance.
+The visualization integrates sequence data with other data types such as geographic information, serology, or host species.
 Nextstrain makes these results, representing our current understanding, publicly available for use by health professionals, epidemiologists, virologists and the public alike.
@@
 Viral pathogens pose an ongoing and ever present danger to human health on a worldwide scale, including epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas, as well as ongoing disease spread such as seasonal influenza.
-The rapid evolution of these viruses allows inference of the evolution and epidemiology of the disease from genomic data.
+The rapid evolution of these viruses allows inference of history and epidemiology of the disease from genomic data.
 Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
@@
 Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
-This rapidity of sequencing must be complemented with rapid methods by which to analyse and interpret results.
+This rapidity of sequencing must be complemented with rapid methods by which to analyse, interpret, and diseminate results.
 
@@
 Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
-These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional metadadata such as genotype or serotype expressed through colourings (figure \ref{nextstrain}).
+These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (figure \ref{nextstrain}).
 This is coupled with a user-interactable time slider to see how the pathogen has evolved and spread over the course of the epidemic.

### 2017-10-10 whitepaper.tex
fixed types

@@
 %Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
-Nextstrain consists of a bioinformatics pipeline for phylodynamics analysis of viral genomes and a to an interactive visualisation platform which presents a real-time view into the evolution and spread of a range of viral pathogens of high human importance.
+Nextstrain consists of database, a bioinformatics pipeline for phylodynamics analysis of viral genomes, and an interactive visualisation platform which presents a real-time view into the evolution and spread of a range of viral pathogens of high human importance.
 The visualization integrates sequence data with other data types such as geographic information, serology, or host species.
@@
 Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
-This rapidity of sequencing must be complemented with rapid methods by which to analyse, interpret, and diseminate results.
+This rapidity of sequencing must be complemented with rapid methods by which to analyse, interpret, and disseminate results.
 

### 2017-10-10 whitepaper.tex
TB's edits

--- 2017-10-10 whitepaper.tex
@@
 Colin Megill\,$^{\text{\sfb 1}}$,
-Sidney M. Bell\,$^{\text{\sfb 1,2}}$,
+Sidney M.~Bell\,$^{\text{\sfb 1,2}}$,
 John Huddleston\,$^{\text{\sfb 1,2}}$,
@@
 %Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
-Nextstrain consists of a bioinformatics pipeline for phylodynamics analysis of viral genomes and a to an interactive visualisation platform which presents a real-time view into the evolution and spread of a range of viral pathogens of high human importance.
+Nextstrain consists of a database of viral genomes, a bioinformatics pipeline for phylodynamics analysis, and an interactive visualisation platform.
+Together these present a real-time view into the evolution and spread of a range of viral pathogens of high public health importance.
 The visualization integrates sequence data with other data types such as geographic information, serology, or host species.
-Nextstrain makes these results, representing our current understanding, publicly available for use by health professionals, epidemiologists, virologists and the public alike.
+Nextstrain compiles our current understanding into a single accessible location, publicly available for use by health professionals, epidemiologists, virologists and the public alike.
 \\
 \textbf{Availability and implementation:}
-All code (predominantly Javascript and Python) is freely available from https://github.com/nextstrain and the web-application is available at www.nextstrain.org.
+All code (predominantly JavaScript and Python) is freely available from \href{https://github.com/nextstrain}{github.com/nextstrain} and the web-application is available at \href{http://nextstrain.org}{nextstrain.org}.
 \\
@@
 
-Viral pathogens pose an ongoing and ever present danger to human health on a worldwide scale, including epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas, as well as ongoing disease spread such as seasonal influenza.
-The rapid evolution of these viruses allows inference of history and epidemiology of the disease from genomic data.
+Viral pathogens pose an ongoing and ever-present danger to global human health, with recent epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas highlighting the risk.
+The rapid evolution of these viruses allows inference of epidemic history from genomic data.
 Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
-Furthermore, the results of analyses are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding.
-We have developed nextstrain to visualise viral outbreaks in as close to real time as possible.
+Furthermore, the results of analyses are often not made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
+We have developed Nextstrain to visualise viral outbreaks in as close to real time as possible.
 % This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
@@
 The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
-Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
-This rapidity of sequencing must be complemented with rapid methods by which to analyse, interpret, and diseminate results.
+Sequencing times and costs are continually dropping, with on-the-ground sequencing being deployed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
+However, this rapidity of sequencing must be complemented with rapid methods by which to analyse, interpret, and disseminate results.
 
 Nextstrain consists of three components:
-``Fauna'' is a collection of python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, GISAID and ViPR, as well as github repositories and other sources of genomic data.
-``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
-This leverages the maximum likelihood phylodynamic analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
-These scripts separate generic core functionality from a light virus specific layer such that they are easily co-opted to different pathogens.
-Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
+\textit{Fauna} is a collection of Python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, GISAID and ViPR, as well as GitHub repositories and other sources of genomic data.
+\textit{Augur} is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
+This leverages the maximum likelihood phylodynamic analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop computer.
+These scripts separate generic core functionality from a light virus-specific layer such that they are easily co-opted to different pathogens.
+Outputs from the pipeline are uploaded online and immediately available through the visualisation software \textit{Auspice}, located at \href{http://nextstrain.org}{nextstrain.org}.
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
-There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.
+There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.
 
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.45\textwidth]{figures/nextstrain}}
-\caption{Currently known genomic epidemiology of Zika virus (\href{nextstrain.org/zika}{nextstrain.org/zika}).
-The main interface consists of three linked panels - a phylogenetic tree, the inferred transmissions projected onto a map, and the genetic diversity across the genome.}
+\centerline{\includegraphics[width=0.48\textwidth]{figures/nextstrain}}
+\caption{Genomic epidemiology of Zika virus as of Oct 2017 (live display at \href{http://nextstrain.org/zika}{nextstrain.org/zika}).
+The main interface consists of three linked panels --- a phylogenetic tree, geographic transmissions, and the genetic diversity across the genome.}
 \label{nextstrain}
@@
 
-\subsection*{Joint Temporal \& Spatial Visualisation}
+\subsection*{Joint temporal and spatial visualisation}
 Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
 Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
-These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (figure \ref{nextstrain}).
+These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (Figure \ref{nextstrain}).
 This is coupled with a user-interactable time slider to see how the pathogen has evolved and spread over the course of the epidemic.
 By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
-This approach both communicates the geographical spread of disease and the stochastic nature of epidemics.
+This approach both communicates the geographical spread of the epidemic alongside the underlying genomic data that supports this geographic reconstruction.
 
-\subsection*{Transmission Chain Analysis}
 Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
@@
 
-\subsection*{Monitoring of viral evolution \& adaptation}
+\subsection*{Monitoring of viral evolution and adaptation}
 Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar-chart of entropy at each position in across the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
@@
 The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
-\textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
+% \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
 Nextstrain allows monitoring of such mutations, such that interventions and control measures may be implemented where appropriate.
 
-\subsection*{A Model for Public Sharing of Data}
-Nextstrain presents a single overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, zika, ebola), all based upon the same underlying bioinformatics architecture.
+\subsection*{A model for public sharing of data}
+Nextstrain presents a single overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, Zika, Ebola), all based upon the same underlying bioinformatics architecture.
 As such, it is well positioned to respond to both the next outbreak and continued surveillance of current ones.
-Analysis of such outbreaks relies on public sharing of data, and nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
-Scientists are justifiably hesitant to share control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
+Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
+Scientists are justifiably hesitant to cede control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
 We do, however provide the ability to download phylogenetic trees, metadata and screenshots, and private metadata can be added by users by simply dragging a CSV file onto the browser.
 We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.
-Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{croucher2015application}.
+Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real-time rather than retrospectively \citep{croucher2015application}.
 
@@
 
-This work has been supported by the OpenSciencePrize to TB and RAN, the ERC through StG-260686 to RAN, and the NSF through DGE-1256082 to SMB.
+This work was supported by the Open Science Prize to TB and RAN, by the NSF through DGE-1256082 to SMB, by the ERC through StG-260686 to RAN and by NIH R35 GM119774-01 to TB.
+TB is a Pew Biomedical Scholar.
 

### 2017-10-10 whitepaper.tex
Minor addition

@@
 Nextstrain presents a single overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, Zika, Ebola), all based upon the same underlying bioinformatics architecture.
-As such, it is well positioned to respond to both the next outbreak and continued surveillance of current ones.
+This same architecture should be fully extendable to bacterial systems in the future.
+It is well positioned to respond to both the next outbreak and continued surveillance of current ones.
 Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
@@
 We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.
-Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real-time rather than retrospectively \citep{croucher2015application}.
+Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential realised by sharing of data in real-time rather than retrospectively \citep{croucher2015application}.
 

### 2017-10-11 whitepaper.tex
john's edits

@@
 These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (Figure \ref{nextstrain}).
-This is coupled with a user-interactable time slider to see how the pathogen has evolved and spread over the course of the epidemic.
-By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
+This is coupled with an interactive time slider to see how the pathogen has evolved and spread over the course of the epidemic.
+By animating the temporal dimension, a high level overview of how the entire outbreak unfolded is quickly gained.
 This approach both communicates the geographical spread of the epidemic alongside the underlying genomic data that supports this geographic reconstruction.
@@
 \subsection*{Monitoring of viral evolution and adaptation}
-Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar-chart of entropy at each position in across the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
-As with all views into data, selecting a position in the genome reveals the selected segregating variant in the phylogeny and the map.
+Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar-chart of entropy at each position in the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
+Selecting a position in the genome with non-zero entropy reveals the disdtribution of the segregating variant in the phylogeny and on the map.
 This allows interrogation of genetic change which may be adaptive or underlying a change in disease dynamics.
@@
 It is well positioned to respond to both the next outbreak and continued surveillance of current ones.
-Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
+Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update as new sequences from a range of public databases and repositories appear.
 Scientists are justifiably hesitant to cede control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
-We do, however provide the ability to download phylogenetic trees, metadata and screenshots, and private metadata can be added by users by simply dragging a CSV file onto the browser.
+Derived data, such as phylogenetic trees, metadata and screenshots may be downloded, and private metadata may be added by users by simply dragging a CSV file onto the browser.
 We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.
@@
 
-\vspace*{-10pt}
+\vspace*{-5pt}
 
@@
 
-\vspace*{-12pt}
+\vspace*{-5pt}
 

### 2017-10-11 whitepaper.tex
james' final edits, incl. title change

@@
 
-\title[Nextstrain]{Nextstrain: real-time tracking of virus evolution}
+\title[Nextstrain]{Nextstrain: real-time tracking of pathogen evolution}
 \author[Hadfield \textit{et~al}.]{
@@
 \textbf{Summary:}
-Understanding the spread and evolution of viral pathogens is important for effective public health measures and surveillance.
+Understanding the spread and evolution of pathogens is important for effective public health measures and surveillance.
 %Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
@@
 Furthermore, the results of analyses are often not made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
-We have developed Nextstrain to visualise viral outbreaks in as close to real time as possible.
+We have developed Nextstrain to visualise outbreaks in as close to real time as possible.
+Whilst currently encompassing a selection of viruses, extension to non-viral pathogens is forthcoming.
 % This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
@@
 Sequencing times and costs are continually dropping, with on-the-ground sequencing being deployed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
-However, this rapidity of sequencing must be complemented with rapid methods by which to analyse, interpret, and disseminate results.
+However, this speed of sequencing must be complemented with rapid methods by which to analyse, interpret, and disseminate results.
 
@@
 This leverages the maximum likelihood phylodynamic analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop computer.
-These scripts separate generic core functionality from a light virus-specific layer such that they are easily co-opted to different pathogens.
+These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
 Outputs from the pipeline are uploaded online and immediately available through the visualisation software \textit{Auspice}, located at \href{http://nextstrain.org}{nextstrain.org}.
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
-There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.
+There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to nearly all outbreaks with readily accessible genomic data.
 
@@
 
-\subsection*{Monitoring of viral evolution and adaptation}
+\subsection*{Monitoring of evolution and adaptation}
 Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar-chart of entropy at each position in the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
-Selecting a position in the genome with non-zero entropy reveals the disdtribution of the segregating variant in the phylogeny and on the map.
+Selecting a position in the genome with non-zero entropy reveals the distribution of the segregating variant in the phylogeny and on the map.
 This allows interrogation of genetic change which may be adaptive or underlying a change in disease dynamics.
@@
 For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
-China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
+For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
 The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
 % \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
-Nextstrain allows monitoring of such mutations, such that interventions and control measures may be implemented where appropriate.
+Nextstrain allows monitoring of experimentally determined mutations, such that interventions and control measures may be implemented where appropriate.
 
 \subsection*{A model for public sharing of data}
-Nextstrain presents a single overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, Zika, Ebola), all based upon the same underlying bioinformatics architecture.
-This same architecture should be fully extendable to bacterial systems in the future.
-It is well positioned to respond to both the next outbreak and continued surveillance of current ones.
+Nextstrain currently presents a single, continuously updated overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, Zika, Ebola), all based upon the same underlying bioinformatics architecture.
+This architecture is well positioned to respond to future outbreaks, be they viral or bacterial.
+
 Analysis of such outbreaks relies on public sharing of data, and Nextstrain has the ability to automatically update as new sequences from a range of public databases and repositories appear.
@@
 
-\vspace*{-5pt}
+\vspace*{-15pt}
 

### 2017-11-22 whitepaper_biorxiv.tex
Include single-column version for bioRxiv

--- 2017-10-11 whitepaper.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
+\documentclass[11pt,oneside,letterpaper]{article}
 
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+% graphicx package, useful for including eps and pdf graphics
+\usepackage{graphicx}
+\DeclareGraphicsExtensions{.pdf,.png,.jpg}
+
+% basic packages
+\usepackage{color}
+\usepackage{parskip}
+\usepackage{float}
+
+\definecolor{red}{rgb}{0.86,0.17,0.17}
+\definecolor{green}{rgb}{0.20,0.50,0.48}
+\usepackage[hidelinks]{hyperref}
+\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=green}
+
+% text layout
+\usepackage{geometry}
+\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
+\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space
+
+% helps to keep figures from being orphaned on a page by themselves
+\renewcommand{\topfraction}{0.85}
+\renewcommand{\textfraction}{0.1}
+
+% bold the 'Figure #' in the caption and separate it with a period
+% Captions will be left justified
+\usepackage[labelfont=bf,labelsep=period,font=small]{caption}
+
+% review layout with double-spacing
+%\usepackage{setspace}
+%\doublespacing
+%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}
+
+% cite package, to clean up citations in the main text. Do not remove.
+% \usepackage{cite}
+\usepackage[numbers]{natbib}
+%\renewcommand\citeleft{(}
+%\renewcommand\citeright{)}
+%\renewcommand\citeform[1]{\textsl{#1}}
+
+% Remove brackets from numbering in list of References
+\renewcommand\refname{References}
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+\usepackage{authblk}
+\renewcommand\Authands{ \& }
+\renewcommand\Authfont{\normalsize \bf}
+\renewcommand\Affilfont{\small \normalfont}
+\makeatletter
+\renewcommand\AB@affilsepx{, \protect\Affilfont}
+\makeatother
+
+%%% TITLE %%%
+\title{\vspace{1.0cm} \Large \bf
+Nextstrain: real-time tracking of pathogen evolution
+}
+
+\author[1,*]{James Hadfield}
+\author[1]{Colin Megill}
+\author[1,2]{Sidney M.~Bell}
+\author[1,2]{John Huddleston}
+\author[1]{Barney Potter}
+\author[1]{Charlton Callender}
+\author[3]{Pavel Sagulenko}
+\author[1,$\S$]{Trevor Bedford}
+\author[3,4,5,$\S$]{Richard A.~Neher}
+
+\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
+\affil[2]{Molecular and Cellular Biology Program, University of Washington, Seattle, WA, USA}
+\affil[3]{Max Planck Institute for Developmental Biology, T\"ubingen, Germany}
+\affil[4]{Biozentrum, University of Basel, Basel, Switzerland}
+\affil[5]{SIB Swiss Institute of Bioinformatics, Basel, Switzerland}
+\affil[*]{To whom correspondence should be addressed}
+\affil[$\S$]{These authors contributed equally to this work}
+
+\date{}
 
 \begin{document}
-\firstpage{1}
 
-\subtitle{Phylogenetics}
+\maketitle
 
-\title[Nextstrain]{Nextstrain: real-time tracking of pathogen evolution}
-\author[Hadfield \textit{et~al}.]{
-James Hadfield\,$^{\text{\sfb 1,}\ast}$,
-Colin Megill\,$^{\text{\sfb 1}}$,
-Sidney M.~Bell\,$^{\text{\sfb 1,2}}$,
-John Huddleston\,$^{\text{\sfb 1,2}}$,
-Barney Potter\,$^{\text{\sfb 1}}$,
-Charlton Callender\,$^{\text{\sfb 1}}$,
-Pavel Sagulenko\,$^{\text{\sfb 3}}$,
-Trevor Bedford\,$^{\text{\sfb 1,}\S}$ and
-Richard A.~Neher\,$^{\text{\sfb 3,4,5}\S}$
-}
-\address{
-$^{\text{\sf 1}}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA,
-$^{\text{\sf 2}}$Molecular and Cellular Biology Program, University of Washington, Seattle, WA, USA and
-$^{\text{\sf 3}}$Max Planck Institute for Developmental Biology, T\"ubingen, Germany.
-$^{\text{\sf 4}}$Biozentrum, University of Basel, Basel, Switzerland.
-$^{\text{\sf 5}}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland.
-}
+%%% ABSTRACT %%%
+\begin{abstract}
 
-\corresp{$^\ast$To whom correspondence should be addressed. $^\S$These authors contributed equally to this work.}
-
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-\abstract{
-\textbf{Summary:}
+\noindent \textbf{Summary:}
 Understanding the spread and evolution of pathogens is important for effective public health measures and surveillance.
-%Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
 Nextstrain consists of a database of viral genomes, a bioinformatics pipeline for phylodynamics analysis, and an interactive visualisation platform.
@@
 \\
-\textbf{Availability and implementation:}
+\noindent \textbf{Availability and implementation:}
 All code (predominantly JavaScript and Python) is freely available from \href{https://github.com/nextstrain}{github.com/nextstrain} and the web-application is available at \href{http://nextstrain.org}{nextstrain.org}.
-\\
-\textbf{Contact:}
-\href{jhadfiel@fredhutch.org}{jhadfiel@fredhutch.org}, \href{tbedford@fredhutch.org}{tbedford@fredhutch.org}, \href{richard.neher@unibas.ch}{richard.neher@unibas.ch}
-\\
-}
 
-\maketitle
+\end{abstract}
 
@@
 Whilst currently encompassing a selection of viruses, extension to non-viral pathogens is forthcoming.
-% This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.
 
@@
 
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.48\textwidth]{figures/nextstrain}}
+\begin{figure}[t]
+\centerline{\includegraphics[width=0.8\textwidth]{figure}}
 \caption{Genomic epidemiology of Zika virus as of Oct 2017 (live display at \href{http://nextstrain.org/zika}{nextstrain.org/zika}).
@@
 
-\vspace*{-15pt}
-
-% please add acknowledgment if applicable.
-%\section*{Acknowledgements}
-%We would like to thank...
-%
-%\vspace*{-12pt}
-
 \section*{Funding}
@@
 
-\vspace*{-15pt}
-
-\bibliographystyle{bioinfo}
+\bibliographystyle{unsrtnat}
 \bibliography{whitepaper}

### 2018-04-16 replyToReviewer.tex
init reply to reviewer

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\begin{document}

Please find our reply to the points raised by reviewer 1.

\begin{quotation}
General comments:
In this well-written and timely Applications Note, Hadfield and colleagues present the Nextstrain platform – a web-based visualization and analysis platform for viral genome phylogenies. This reviewer has used the site on many occasions to explore viral genomic data, and can attest to the fact that Nextstrain is an incredibly useful resource to the viral genomics and public health communities. Its thoughtful, intuitive interface makes the tool easy to use, even for individuals with no formal bioinformatics training. Nextstrain has played a role in a number of recent outbreaks of public health importance, from Ebola to Zika, and is an excellent resource for annual surveillance of influenza viruses. The platform's commitment to open science is exceptional, and demonstrates to the viral genomics community how powerful data-sharing can be.

\paragraph{}
This is an excellent Applications Note describing a fantastic tool, and I have no hesitations in recommending the manuscript for publication. I have no major revisions to suggest, and only very minor style suggestions that the authors may take or leave:

\end{quotation}

1. Page 1, Line 39: "ongoing and ever-present" is somewhat redundant; I'd just go with "ever-present"
2. Page 1, Line 40: perhaps change "recent epidemics" to "recent events" so "epidemics" isn't repeated 3x in the sentence.

3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.

4. Page 1, Line 43 (column 2): could references to the three databases (papers or URLs) be included here?

5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of "nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?

6. Page 2, Line 51: it should be noted that the transmission inference depends on how well the outbreak in question was sampled - if data is missing, important transmission links could be missed.

7. Page 2, Line 20 (column 20): is there an example of one of these mutations that could be mentioned here?

There is a fair amount of passive voice in the manuscript - popping the text into readable.io to find these instances and changing the construction to make them active will make the writing a little more lively (and it usually frees up some words in the word count).

\end{document}

### 2018-04-16 replyToReviewer.tex
points 1 & 2

@@
 
+We thank the reviewer for these kind words!
+
+\begin{quote}
 1. Page 1, Line 39: "ongoing and ever-present" is somewhat redundant; I'd just go with "ever-present"
+
 2. Page 1, Line 40: perhaps change "recent epidemics" to "recent events" so "epidemics" isn't repeated 3x in the sentence.
+\end{quote}
 
+We have changed the wording of the manuscript to address these points.
+
+\begin{quote}
 3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.
+\end{quote}
 

### 2018-04-16 whitepaper.tex
changes for submission
points 1 & 2

@@
 Understanding the spread and evolution of pathogens is important for effective public health measures and surveillance.
-%Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
 Nextstrain consists of a database of viral genomes, a bioinformatics pipeline for phylodynamics analysis, and an interactive visualisation platform.
@@
 
-Viral pathogens pose an ongoing and ever-present danger to global human health, with recent epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas highlighting the risk.
+Viral pathogens pose an ever-present danger to global human health, with recent events such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas highlighting the risk.
 The rapid evolution of these viruses allows inference of epidemic history from genomic data.
@@
 \begin{figure}[!tpb]
-\centerline{\includegraphics[width=0.48\textwidth]{figures/nextstrain}}
+\centerline{\includegraphics[width=0.48\textwidth]{figure}}
 \caption{Genomic epidemiology of Zika virus as of Oct 2017 (live display at \href{http://nextstrain.org/zika}{nextstrain.org/zika}).
@@
 %\section*{Acknowledgements}
-%We would like to thank...
-%
 %\vspace*{-12pt}

### 2018-04-17 replyToReviewer.tex
reviewer points 3 & 4

@@
 
-Please find our reply to the points raised by reviewer 1.
-
-\begin{quotation}
-General comments:
-In this well-written and timely Applications Note, Hadfield and colleagues present the Nextstrain platform – a web-based visualization and analysis platform for viral genome phylogenies.
-This reviewer has used the site on many occasions to explore viral genomic data, and can attest to the fact that Nextstrain is an incredibly useful resource to the viral genomics and public health communities.
-Its thoughtful, intuitive interface makes the tool easy to use, even for individuals with no formal bioinformatics training.
-Nextstrain has played a role in a number of recent outbreaks of public health importance, from Ebola to Zika, and is an excellent resource for annual surveillance of influenza viruses.
-The platform's commitment to open science is exceptional, and demonstrates to the viral genomics community how powerful data-sharing can be.
-
-\paragraph{}
-This is an excellent Applications Note describing a fantastic tool, and I have no hesitations in recommending the manuscript for publication.
-I have no major revisions to suggest, and only very minor style suggestions that the authors may take or leave:
-
-\end{quotation}
-
-We thank the reviewer for these kind words!
+We thank the reviewer for their kind words, and have addressed all of the points (detailed below).
 
@@
 \end{quote}
+We have reworded this section to include a reference to ``Viral Phylodynamics'', Voltz \textit{et al.} 2013.
 
+\begin{quote}
 4. Page 1, Line 43 (column 2): could references to the three databases (papers or URLs) be included here?
+\end{quote}
+We have added URLs to the text.
 
+\begin{quote}
 5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of "nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?
+\end{quote}
 
+\begin{quote}
 6. Page 2, Line 51: it should be noted that the transmission inference depends on how well the outbreak in question was sampled - if data is missing, important transmission links could be missed.
+\end{quote}
 
+\begin{quote}
 7. Page 2, Line 20 (column 20): is there an example of one of these mutations that could be mentioned here?
+\end{quote}
 
+\begin{quote}
 There is a fair amount of passive voice in the manuscript - popping the text into readable.io to find these instances and changing the construction to make them active will make the writing a little more lively (and it usually frees up some words in the word count).
 
+\end{quote}
+
 \end{document}

### 2018-04-17 whitepaper.tex
reviewer points 3 & 4

@@
 Nextstrain consists of three components:
-\textit{Fauna} is a collection of Python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, GISAID and ViPR, as well as GitHub repositories and other sources of genomic data.
-\textit{Augur} is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
-This leverages the maximum likelihood phylodynamic analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop computer.
+\textit{Fauna} is a collection of Python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
+\textit{Augur} is a suite of tools to perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
+This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
 These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.

### 2018-04-17 replyToReviewer.tex
point 5

@@
 \end{quote}
+There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.''
 

### 2018-04-17 whitepaper.tex
point 5

@@
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
-There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to nearly all outbreaks with readily accessible genomic data.
+There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.
 

### 2018-04-17 replyToReviewer.tex
point 6

@@
 \end{quote}
+This is a very important point, and we have added the following sentence to the manuscript: ``It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.''
 
@@
 \end{quote}
+We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificit'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org
 

### 2018-04-17 whitepaper.tex
point 6

@@
 Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
+It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.
 Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
@@
 For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
-The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
+The threat of mutations which facilitate human to human transmission is of extreme concern (e.g. see \citep{deVries2017}), as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
 % \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.

### 2018-04-17 whitepaper.tex
remove 11/12 usages of the passive voice

--- 2018-04-17 whitepaper.tex
@@
 The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
-Sequencing times and costs are continually dropping, with on-the-ground sequencing being deployed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
-However, this speed of sequencing must be complemented with rapid methods by which to analyse, interpret, and disseminate results.
+Sequencing times and costs are continually dropping, with on-the-ground sequencing used during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
+Rapid methods by which to analyse, interpret, and disseminate results must complement this sepeed of sequencing.
 
@@
 These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
-Outputs from the pipeline are uploaded online and immediately available through the visualisation software \textit{Auspice}, located at \href{http://nextstrain.org}{nextstrain.org}.
+Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
@@
 Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
-Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
-These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (Figure \ref{nextstrain}).
+Multiple views into different facets of the data are presented and remain in sync as one interacts with the data.
+This allows simultaneous interrogation of phylogenetic and spatial relationships, with additional data such as genotype or serotype expressed through colourings (Figure \ref{nextstrain}).
 This is coupled with an interactive time slider to see how the pathogen has evolved and spread over the course of the epidemic.
@@
 Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
-Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
-It is important to make known the confidence of any such reconstruction, and we do this in two ways: by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
+Internal node colours indicate inferred transmissions on the tree, which are also drawn as links between demes on the map.
+It is important to make known the uncertainty of such reconstructions.
+The confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
+Despite this, sampling bias can obscure transmission links, and in cases of extreme bias we have chosen not to display the inferred states.
 
@@
 For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
-For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
-The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
-% \textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
-Nextstrain allows monitoring of experimentally determined mutations, such that interventions and control measures may be implemented where appropriate.
+For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years.
+Despite no known cases of human to human transmission, the high mortality rate of $~30\%$ \citep{li2014epidemiology} makes the threat of mutations which facilitate human to human transmission of extreme concern (e.g.
+see \citep{devries2017three}).
+Nextstrain allows monitoring of experimentally determined mutations, informing appropriate interventions and control measures.
 
@@
 Scientists are justifiably hesitant to cede control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
-Derived data, such as phylogenetic trees, metadata and screenshots may be downloded, and private metadata may be added by users by simply dragging a CSV file onto the browser.
+Derived data, such as phylogenetic trees, metadata and screenshots are available, and private metadata added via CSV files.
 We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, thereby encouraging collaboration between scientists.

### 2018-04-17 whitepaper.tex
remove fauna / augur / auspice terminology

@@
 
-Nextstrain consists of three components:
-\textit{Fauna} is a collection of Python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
-\textit{Augur} is a suite of tools to perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
+Nextstrain consists of data curation, analysis and visualisation components:
+Python scripts maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
+A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
 This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.

### 2018-04-17 replyToReviewer.tex
remove 11/12 usages of the passive voice
simplify sentences

@@
 \end{quote}
-
 We have changed the wording of the manuscript to address these points.
@@
 There is a fair amount of passive voice in the manuscript - popping the text into readable.io to find these instances and changing the construction to make them active will make the writing a little more lively (and it usually frees up some words in the word count).
+\end{quote}
+Thankyou for alerting me us to this website. We have removed 11/12 usages of the passive voice, as well as simplifying sentences with too many syllables where possible.
 
-\end{quote}
+\paragraph{}
+Sincerely,
+
+James Hadfield
 

### 2018-04-17 whitepaper.tex
simplify sentences

@@
 
-Viral pathogens pose an ever-present danger to global human health, with recent events such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas highlighting the risk.
+Viral pathogens pose an ever-present danger to global human health, highlighted by recent events such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas.
 The rapid evolution of these viruses allows inference of epidemic history from genomic data.
 Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
-Furthermore, the results of analyses are often not made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
+Furthermore, the results of analyses are rarely made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
 We have developed Nextstrain to visualise outbreaks in as close to real time as possible.
@@
 The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
-Sequencing times and costs are continually dropping, with on-the-ground sequencing used during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
+Sequencing times and costs are continually dropping, with on-the-ground sequencing used during recent epidemics \citep{quick2016real,faria2017epidemic}.
 Rapid methods by which to analyse, interpret, and disseminate results must complement this sepeed of sequencing.
@@
 It is important to make known the uncertainty of such reconstructions.
-The confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
-Despite this, sampling bias can obscure transmission links, and in cases of extreme bias we have chosen not to display the inferred states.
+Confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
+Sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states.
 
@@
 For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years.
-Despite no known cases of human to human transmission, the high mortality rate of $~30\%$ \citep{li2014epidemiology} makes the threat of mutations which facilitate human to human transmission of extreme concern (e.g.
-see \citep{devries2017three}).
+Despite no known cases of human to human transmission, the high mortality rate of $~30\%$ \citep{li2014epidemiology} makes the threat of mutations which facilitate human to human transmission of extreme concern (see \citep{devries2017three}).
 Nextstrain allows monitoring of experimentally determined mutations, informing appropriate interventions and control measures.

### 2018-04-17 replyToReviewer.tex
insignificant changes in response

@@
 \begin{quote}
-1. Page 1, Line 39: "ongoing and ever-present" is somewhat redundant; I'd just go with "ever-present"
+1. Page 1, Line 39: ``ongoing and ever-present" is somewhat redundant; I'd just go with ``ever-present"
 
-2. Page 1, Line 40: perhaps change "recent epidemics" to "recent events" so "epidemics" isn't repeated 3x in the sentence.
+2. Page 1, Line 40: perhaps change ``recent epidemics" to ``recent events" so ``epidemics" isn't repeated 3x in the sentence.
 \end{quote}
@@
 \begin{quote}
-5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of "nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?
+5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of ``nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?
 \end{quote}
-There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.''
+There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.''
 
@@
 \end{quote}
-We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificit'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org
+We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificity'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org
 

### 2018-04-17 whitepaper.tex
insignificant changes in response

@@
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
-There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.
+There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.
 

### 2018-04-17 whitepaper.tex
small edits

@@
 Sequencing times and costs are continually dropping, with on-the-ground sequencing used during recent epidemics \citep{quick2016real,faria2017epidemic}.
-Rapid methods by which to analyse, interpret, and disseminate results must complement this sepeed of sequencing.
+Rapid methods by which to analyse, interpret, and disseminate results must complement this speed of sequencing.
 
 Nextstrain consists of data curation, analysis and visualisation components:
-Python scripts maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
+Python scripts maintain a database of available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
 A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
 This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
-These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
+These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily adapted to different pathogens.
 Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
@@
 Confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
-Sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states.
+Sampling bias and lack of data can obscure transmission links, and in certain cases we have chosen not to display the inferred states.
 

### 2018-04-18 replyToReviewer.tex
Slightly prettify response to reviewer
No textual changes here

--- 2018-04-17 replyToReviewer.tex
@@
-\documentclass[12pt,a4paper]{article}
-\usepackage[utf8]{inputenc}
+\documentclass[11pt,oneside,letterpaper]{article}
+
+% graphicx package, useful for including eps and pdf graphics
+\usepackage{graphicx}
+\DeclareGraphicsExtensions{.pdf,.png,.jpg}
+
+% basic packages
+\usepackage{color}
+\usepackage{parskip}
+\usepackage{float}
+\usepackage{hyperref}
+
+% text layout
+\usepackage{geometry}
+\geometry{textwidth=16.25cm} % 15.25cm for single-space, 16.25cm for double-space
+\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space
+
+% helps to keep figures from being orphaned on a page by themselves
+\renewcommand{\topfraction}{0.85}
+\renewcommand{\textfraction}{0.1}
+
+% bold the 'Figure #' in the caption and separate it with a period
+% Captions will be left justified
+\usepackage[labelfont=bf,labelsep=period,font=small]{caption}
+
+% review layout with double-spacing
+%\usepackage{setspace}
+%\doublespacing
+%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+%\renewcommand\citeleft{(}
+%\renewcommand\citeright{)}
+%\renewcommand\citeform[1]{\textsl{#1}}
+
+% Remove brackets from numbering in list of References
+\renewcommand\refname{\large References}
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+\usepackage{authblk}
+\renewcommand\Authands{ \& }
+\renewcommand\Authfont{\normalsize \bf}
+\renewcommand\Affilfont{\small \normalfont}
+\makeatletter
+\renewcommand\AB@affilsepx{, \protect\Affilfont}
+\makeatother
+
+% notation
 \usepackage{amsmath}
-\usepackage{amsfonts}
 \usepackage{amssymb}
+
 \begin{document}
+
+\newgeometry{top=4cm}
+
+Dear Bioinformatics editorial board,
+
+Thank you for conducting a review of our manuscript entitled ``Nextstrain: real-time tracking of pathogen evolution''.
+We have responded to the reviewer criticisms and believe the manuscript is now suitable for publication in Bioinformatics.
+
+Point-by-point responses follow, as well as a PDF showing revisions that have been made since the last submission.
+
+Sincerely,\\
+James Hadfield
+
+\restoregeometry
+
+\newpage
+
+\section*{Reviewer responses}
+
+Original reviewer criticisms are in plain text.  Our responses follow in \textbf{bold}.
 
@@
 
-\begin{quote}
-1. Page 1, Line 39: "ongoing and ever-present" is somewhat redundant; I'd just go with "ever-present"
+1. Page 1, Line 39: ``ongoing and ever-present'' is somewhat redundant; I'd just go with ``ever-present''
 
-2. Page 1, Line 40: perhaps change "recent epidemics" to "recent events" so "epidemics" isn't repeated 3x in the sentence.
-\end{quote}
+2. Page 1, Line 40: perhaps change ``recent epidemics'' to ``recent events'' so ``epidemics'' isn't repeated 3x in the sentence.
 
-We have changed the wording of the manuscript to address these points.
+\textbf{We have changed the wording of the manuscript to address these points.}
 
-\begin{quote}
 3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.
-\end{quote}
-We have reworded this section to include a reference to ``Viral Phylodynamics'', Voltz \textit{et al.} 2013.
 
-\begin{quote}
+\textbf{We have reworded this section to include a reference to ``Viral Phylodynamics'', Voltz \textit{et al.} 2013.}
+
 4. Page 1, Line 43 (column 2): could references to the three databases (papers or URLs) be included here?
-\end{quote}
-We have added URLs to the text.
 
-\begin{quote}
-5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of "nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?
-\end{quote}
-There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or slow mutation rate to confound phylogenetic signal.''
+\textbf{We have added URLs to the text.}
 
-\begin{quote}
+5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of ``nearly all outbreaks'' -- which ones aren't amenable to Nextstrain-ification?
+
+\textbf{There are indeed pathogens for which the general approach outlined in this paper is insufficient, and so we have changed the text to read ``Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.''}
+
 6. Page 2, Line 51: it should be noted that the transmission inference depends on how well the outbreak in question was sampled - if data is missing, important transmission links could be missed.
-\end{quote}
-This is a very important point, and we have added the following sentence to the manuscript: ``It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.''
 
-\begin{quote}
+\textbf{This is a very important point, and we have added the following sentence to the manuscript: ``It is important to note that sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states due to this bias.''}
+
 7. Page 2, Line 20 (column 20): is there an example of one of these mutations that could be mentioned here?
-\end{quote}
-We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificit'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org
 
-\begin{quote}
-There is a fair amount of passive voice in the manuscript - popping the text into readable.io to find these instances and changing the construction to make them active will make the writing a little more lively (and it usually frees up some words in the word count).
+\textbf{We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificity'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org}
 
-\end{quote}
+There is a fair amount of passive voice in the manuscript -- popping the text into readable.io to find these instances and changing the construction to make them active will make the writing a little more lively (and it usually frees up some words in the word count).
+
+\textbf{Thank you for alerting us to this website. We have removed 11/12 usages of the passive voice, as well as simplifying sentences with too many syllables where possible.}
 

### 2018-04-18 replyToReviewer.tex
Revision pass

@@
 
-\textbf{We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificity'' (de Vries \textit{et al.}), however have not listed these mutations as we detect none in the (consensus) sequences shown on nextstrain.org}
+\textbf{We have added a reference to the 2017 publication ``Three mutations switch H7N9 influenza to human-type receptor specificity'' (de Vries \textit{et al.}).}
 

### 2018-04-18 whitepaper.tex
Revision pass

@@
 The visualization integrates sequence data with other data types such as geographic information, serology, or host species.
-Nextstrain compiles our current understanding into a single accessible location, publicly available for use by health professionals, epidemiologists, virologists and the public alike.
+Nextstrain compiles our current understanding into a single accessible location, open to health professionals, epidemiologists, virologists and the public alike.
 \\
@@
 
-Viral pathogens pose an ever-present danger to global human health, highlighted by recent events such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas.
+Viral pathogens pose an ever-present danger to global human health, highlighted by recent events such as the West African Ebola epidemic and the ongoing Zika epidemic in the Americas.
 The rapid evolution of these viruses allows inference of epidemic history from genomic data.
-Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
-Furthermore, the results of analyses are rarely made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
+Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results \citep{pybus2013evolutionary}.
+Furthermore, the results of analyses are rarely made available to the public or health bodies until after publication, which may be too late to aid understanding or effect change in policy.
 We have developed Nextstrain to visualise outbreaks in as close to real time as possible.
@@
 Python scripts maintain a database of available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
-A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
-This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
+A suite of tools perform phylodynamic analysis \citep{volz2013viral}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
+This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2018treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
 These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily adapted to different pathogens.
-Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
+Visualisation is available through \href{http://nextstrain.org}{nextstrain.org}.
 This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
@@
 
-Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
-Internal node colours indicate inferred transmissions on the tree, which are also drawn as links between demes on the map.
+Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with inferred probability distribution of ancestral state at each node.
+Internal node colours indicate ancestral state and shifts in ancestral state are drawn as links between demes on the map.
 It is important to make known the uncertainty of such reconstructions.
@@
 \subsection*{Monitoring of evolution and adaptation}
-Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar-chart of entropy at each position in the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
+Nextstrain tracks and reconstructs mutations across the tree and displays this information as a bar chart of entropy at each position in the genome, as well as showing the mutations inferred to occur on each branch by hovering over the tree.
 Selecting a position in the genome with non-zero entropy reveals the distribution of the segregating variant in the phylogeny and on the map.
@@
 For instance, China has experienced seasonal epidemics of influenza A/H7N9 over the past five years.
-Despite no known cases of human to human transmission, the high mortality rate of $~30\%$ \citep{li2014epidemiology} makes the threat of mutations which facilitate human to human transmission of extreme concern (see \citep{devries2017three}).
-Nextstrain allows monitoring of experimentally determined mutations, informing appropriate interventions and control measures.
+Despite no known human-to human-transmission events, the high mortality rate of $~30\%$ \citep{li2014epidemiology} makes the threat of mutations which facilitate human-to-human transmission of extreme concern.
+For example, mutations identified by \citet{devries2017three} are readily visible at \href{http://nextstrain.org/avian/h7n9}{nextstrain.org/avian/h7n9}.
+Continual monitoring of such putatively adaptive mutations is critical.
 
 \subsection*{A model for public sharing of data}
-Nextstrain currently presents a single, continuously updated overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, Zika, Ebola), all based upon the same underlying bioinformatics architecture.
+Nextstrain presents a single, continuously updated overview of both endemic viral disease (seasonal influenza, dengue) as well as emergent viral outbreaks (avian influenza, Zika, Ebola), all based upon the same underlying bioinformatics architecture.
 This architecture is well positioned to respond to future outbreaks, be they viral or bacterial.

### 2018-04-18 whitepaper.tex
Small fix

@@
 
-Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with inferred probability distribution of ancestral state at each node.
+Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with inferred probability distributions of ancestral state at each node.
 Internal node colours indicate ancestral state and shifts in ancestral state are drawn as links between demes on the map.
-It is important to make known the uncertainty of such reconstructions.
 Confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
